---
title: "DS202 Final Project"
output: github_document
author: 'Alex Mears'
---

```{r, include = FALSE}
library(tidyverse)
library(scales)
library(rvest)

# Use for local editing
df <- read.csv("career_explorer_occupational_level_data_1192_rows.csv")

# Use for uploading
# df <- read.csv("https://raw.githubusercontent.com/ammears03/DS2020_Final/refs/heads/main/career_explorer_occupational_level_data_1192_rows.csv")
```

# Misinformation: The Most Dangerous Occupational Hazard

## Introduction

For this project, I will be examining the [**Career Explorer Occupational Level Data**](https://data.iowa.gov/catalog/dataset/1192) data set to get some insights into the job market in Iowa. This data set is created and managed by the Iowa Workforce Development agency, and it is updated every year. So the data is comprehensive, clean, and up to date.

This data set compiles information about different jobs across Iowa. It is incredibly large, at 40048 rows and 91 columns. The data set includes information about current job markets, predicted future changes in those markets, education requirements, desired skills, and much more. For a full breakdown of each column, check out the link above, and click on the 'Columns' tab. Since there are so many, I will not go into detail on each one, but I will describe below the ones I most often use in my research. Otherwise I will describe columns as I use them.

1.  geography - County location of data point.
2.  occupational_title - Title of occupation.
3.  estimated_employment - Estimated number of current jobs for occupation in geographical location.
4.  projected_employment - Projected number of future jobs for occupation in geographical location.
5.  average_annual_salary - Average annual salary of occupation in geographical location.
6.  typical_education_requirement - Typical education requirement for occupation.
7.  on_the_job_training_requirement - Typical on the job training requirement for occupation.
8.  skill_def - Short descriptive definition of the top skill for occupation.
9.  skill_type - Type of skill that skill_def is categorized by.

## Initial Overview

Firstly, I want to take a look at some of the more 'basic' explorations.

### How many different jobs are there?

```{r}
cat('There are', 
    df |> select(occupational_title) |> unique()|> nrow(),
    'unique occupational titles')
```

### How many years does this data set cover?

The data set has a few different columns related to years. `estimated_employment_year` gives the year that `estimated_employment` was estimated. `projected_employment_year` gives the year for the `projected_employment`. `wage_year` identifies the year when the wage data was collected.

```{r}
df |> select(wage_year) |> unique()
```

```{r}
df |> select(estimated_employment_year) |> unique()
```

```{r}
df |> select(projected_employment_year) |> unique()
```

For all three columns, they each have one unique value. For all jobs, the `estimated_employment` was collected in 2022, `projected_employment` is for 2032, and wages were collected in 2024.

### What does the average salary distribution look like?

For these next few, I will just look at rows with `geography == 'State of Iowa'`. Since this is just a basic exploration of the data, I don't want to get too in the weeds quite yet. In my more advanced exploration I will be exploring the data as it relates to counties in Iowa.

```{r}
dfia <- df |> filter(geography == 'State of Iowa')
```

```{r, warning=FALSE}
dfia |> select(average_annual_salary) |> ggplot(aes(x = average_annual_salary)) + geom_histogram(bins = 100) + scale_x_continuous(labels = label_comma(), n.breaks = 10)+labs(x='Average annual salary', y='Number of jobs', title = 'State of Iowa Annual Salary Histogram')
```

```{r}
dfia |> select(average_annual_salary) |> summary()
```

From this histogram and the summary we see that most jobs fall within the 25,000 - 80,000 range, with a few large outliers.

### What jobs have the most employment? Which have the least?

```{r}
search_max_min <- dfia |> select(occupational_title, estimated_employment) 
search_max_min[which.max(search_max_min$estimated_employment), ]

search_max_min[which.min(search_max_min$estimated_employment), ]
```

Understandably, the job with the highest employment is `Farmers, Ranchers, & Other Agricultural Manager` with 86,365 jobs. The job with the lowest employment is `Motion Picture Projectionist`, with only 85 jobs.

### What are the most common top ranked skills?

```{r}
dfia |> select(skill_def, skill_type) |> 
  ggplot(aes(x=fct_rev(fct_infreq(skill_def)), fill = skill_type)) + 
  geom_bar() + coord_flip() +
  geom_text(
    stat = 'count',
    aes(label = after_stat(count)),
    vjust = 0.4,
    hjust = 1.1,
    size = 3
  )+labs(x='Skill', y='Number of jobs', title = 'Distribution of Top Skills')
```

The most frequent top skill is `Active Listening` by a large margin. The next highest skill is Critical Thinking, with 134 less jobs listing it as the top skill. Most interestingly though, is that 70 jobs didn't have a skill listed. Those jobs being:

```{r}
dfia |> filter(top_skill_code == 'N.A.') |> select(occupational_title)
```

When looking at the jobs without listed skills, we see some pretty popular roles listed. For example, `Social Worker` is a job without a listed skill, but I would think that something like `Active Listening` would absolutely be one of the top skills for that role.

Ironically enough, one of the jobs without skill listed is `Data Scientist`.

### What are typical education requirements?

```{r}
dfia |> select(typical_education_requirement) |> ggplot(aes(fct_rev(fct_infreq(typical_education_requirement)))) + geom_bar() + coord_flip()+labs(x='Education Requirement', y='Unique Jobs', title = 'Minimum Education Requirement')
```

We see that `High School Diploma or Equivalent` is good enough to get you the majority of jobs in the state. Following that is `Bachelor's Degree`, and then `No Formal Educational Credential`. The rest of the educational levels have much fewer jobs, with the least being `Some College, No Degree`.

## Projected Future Exploration

One of the features of this data set is the projected growth data. The `projected_employment` column provides an estimation for the amount of jobs in 2032. Using this column, I want to explore what the future job market is expected to look like.

Initially, let's just explore what jobs are expected to have the most growth.

```{r}
# What jobs have the most projected growth/employment? What skills, education, training is required. 
dffut <- df |> mutate(deljobs = projected_employment - estimated_employment) |> slice_max(order_by = deljobs, n = 100) # Creating df with 100 highest growing jobs (jobs with most projected new openings) 
dffut |> select(geography,occupational_title,deljobs) |> head(10)
```

Just looking at the top 10, we see a wide variety of jobs. Notably 5 of the top 10 are in the healthcare/hospitality sector. The rest of the top 10 are spread out across more 'blue collar' sectors.

### What are the projected top skills, education, and training for the future?

```{r}
dffut|> select(skill_def, skill_type) |>    ggplot(aes(x=fct_rev(fct_infreq(skill_def)), fill = skill_type)) +    geom_bar() + coord_flip() +   geom_text(     stat = 'count',     aes(label = after_stat(count)),     vjust = 0.4,     hjust = 1.1,     size = 3   )+labs(x='Skill', y='Unique Jobs', title = 'Future Distribution of Top Skills')
```

Here we can see an overwhelming need for `Active Listening` in the future. This isn't too surprising, since earlier we found `Active Listening` to be the most frequent top skill for the present day.

```{r}
dffut |> select(typical_education_requirement) |> ggplot(aes(x=fct_rev(fct_infreq(typical_education_requirement)))) + geom_bar() + coord_flip()+labs(x='Education Requirement', y='Unique Jobs', title = 'Future Minimum Education Requirement by Unique Jobs')
```

Again, only a `High School Diploma or Equivalent` is needed for most of the highest growing jobs, which was true for present day jobs. This chart only counts the number of occupations with these education requirements. This may skew the data a bit, since we saw 5 of the top 10 jobs were in healthcare, which requires more than a High School Diploma. What does it look like if we compare it to the number of projected jobs instead?

```{r,fig.width = 8}
dffut |> summarize(tot_jobs = sum(projected_employment), .by=typical_education_requirement) |> ggplot(aes(x=reorder(typical_education_requirement,tot_jobs), y=tot_jobs))+geom_col()+coord_flip()+labs(x='Education Requirement', y='Total Number of Jobs', title = 'Future Minimum Education Requirement by Total Number of Jobs')+scale_y_continuous(labels = label_number())
```

We still see `High School Diploma or Equivalent` as the largest requirement, but `Postsecondary Non-Degree Award` has crept up into the second spot.

```{r, fig.width=8}
dffut |> select(on_the_job_training_requirement) |> ggplot(aes(x=on_the_job_training_requirement)) + geom_bar() + coord_flip()+labs(x='On the Job Training Requirement', y='Unique Jobs', title = 'Future On the Job Training Requirements by Unique Jobs')
```

This graph clearly shows that most of the highest growing jobs require `Short-Term On-the-Job Training`. This follows from our previous graphs. If most jobs only require a `High School Diploma or Equivalent`, and no exceptionally technical skills, the jobs would require some training, but not a large amount.

Again, this graph charts the training requirements to the number of occupations, not the number of projected jobs.

```{r, fig.width=8}
dffut |> summarize(tot_jobs = sum(projected_employment), .by = on_the_job_training_requirement)|> ggplot(aes(x=on_the_job_training_requirement, y=tot_jobs)) + geom_col() + coord_flip()+labs(x='On the Job Training Requirement', y='Total Future Jobs', title = 'Future On the Job Training Requirements by Total Jobs')+scale_y_continuous(labels = label_number())
```

Modifying the graph to look at the total projected jobs, there is virtually no difference.

In summary, the Iowa job market is predicted to grow most in jobs with lower education requirements, focusing on short term training, with lower technical skill requirements.

## Personal Market Analysis

The primary goal of this project is to explore the Iowa job market in a way that gives me a good insight into my future work environments. To this end, I want to specifically look at two factors, location and occupation.

### Part 1: Location

I am a person who is naturally drawn to more populated areas, so I want to explore what the markets look like in larger counties. For this exploration, I will be looking at counties with a population over 100,000.

```{r}
# Scraping wikipedia for county data

counties <- "https://en.wikipedia.org/wiki/List_of_counties_in_Iowa" |> read_html() |> html_table(fill=T) # Scraping wiki page for tables

IaCount <- counties[[2]] |> select(County, `Population[13]`, `Area[4]`) # Selecting table and useful columns

IaCount <- IaCount |> rename(Population = `Population[13]`, Area = `Area[4]`) # Renaming columns

IaCount$Population <- parse_number(IaCount$Population) # Setting `Population` to numeric values

IaCount$County <- IaCount$County |> str_replace(' County', '') # Getting rid of the word County, to match with other df
```

```{r}
bigPop <- IaCount |> filter(Population >= 100000) # Creating df with just counties with population >= 100,000
```

```{r}
dflarge <- df |> filter(
  geography %in% bigPop$County
) # Filtering our dataset to only include the large counties
```

Initially, I just want to explore what the general job market looks like. To start, let's look at annual salary.

```{r, warning=F}
dflarge |> select(average_annual_salary) |> ggplot(aes(x = average_annual_salary)) + geom_histogram(bins = 100) + scale_x_continuous(labels = label_comma(), n.breaks = 10)+labs(x='Annual Salary', y='Number of Jobs', title = 'Annual Salary Histogram of Most Populus Counties')
```

```{r}
dflarge |> select(average_annual_salary) |> summary()
```

The histogram and summary paint a very similar picture to our previous examination of the entire state of Iowa. However, our initial look used the data with `geography == State of Iowa`. What does the histogram and summary look like if we open it up to the entire data set?

```{r, warning=F}
df |> select(average_annual_salary) |> ggplot(aes(x = average_annual_salary)) + geom_histogram(bins = 100) + scale_x_continuous(labels = label_comma(), n.breaks = 10)+labs(x='Annual Salary', y='Number of Jobs', title = 'Annual Salary Histogram of all Counties')
```

```{r}
df |> select(average_annual_salary) |> summary()
```

Again, it is nearly identical. So we can say that the salary expectations and tendencies do not change when comparing the overall state to the larger counties. Just for fun, what do these graphs look like when looking at the smallest counties?

```{r, warning=F}
smallPop <- IaCount |> filter(Population <= 10000)

dfsmall <- df |> filter(geography %in% smallPop$County)

dfsmall |> select(average_annual_salary) |> ggplot(aes(x = average_annual_salary)) + geom_histogram(bins = 100) + scale_x_continuous(labels = label_comma(), n.breaks = 10)+labs(x='Annual Salary', y='Number of Jobs', title = 'Annual Salary Histogram of Least Populus Counties')

dfsmall |> select(average_annual_salary) |> summary()
```

While the smaller counties are slightly shifted to the left, it is basically negligible. Overall, we can see that the market doesn't change significantly depending on which counties we are looking at, at least when looking at annual salary.

Getting more specific, what does the market look like in terms of individual occupations in the larger counties?

```{r}
cursaljob <- dflarge |> summarize(
  .by=occupational_title,
  avg_salary = mean(average_annual_salary,na.rm=T),
  number_of_jobs = sum(estimated_employment)
)
cursaljob |> arrange(desc(avg_salary)) |> head(10)

cursaljob |> arrange(desc(number_of_jobs)) |> head(10)
```

From looking at the highest paid occupations, we see a lot of healthcare and engineering roles. The only surprising occupation to me is `Musician & Singer`, which is surely skewed by a few large artists.

Looking at most number of jobs, again we see a lot of high volume, low education and training requirement, hands on jobs. Interestingly though, some of these jobs have a fairly high average salary, with `Farmers, Ranchers, & Other Agricultrual Manager` nearing 6 figures.

The ultimate takeaway from this analysis is that the job market as a whole, in Iowa, doesn't really change when looking at different locations. For me, this means that I need to focus more on occupation, and not concern myself too much about location.

### Part 2: Occupation

```{r eval=FALSE, include=FALSE}
df |> select(occupational_title) |> unique() |> View()
# This was used to look through the different occupations, in order to select which ones I wanted to explore further.
```

After looking through the different occupations, I selected 10 occupations that either I am interested in, currently have, or have had in the past. I wanted to include some jobs that I have had, to compare and contrast with the jobs I am looking to have in the future.

```{r}
# Filtering by occupations that I am interested in or have done in the past
my_occupations <- 
  c('Mathematical Science Teacher, Postsecondary','Secondary School Teacher, Except Special & Career/Technical Education','Art, Drama, & Music Teacher, Postsecondary','Teacher and Instructor, All Other','Data Scientist','Actuary','Photographer','Bartender','Tutor','Statistician')
dfmyocc <- df |> filter(occupational_title %in% my_occupations)
dfmyocc$occupational_title[dfmyocc$occupational_title == 'Secondary School Teacher, Except Special & Career/Technical Education'] <- 'Secondary School Teacher' # Shortens job title for nicer graphs
```

Initially, lets just see how many jobs are predicted to be available for each of the occupations in the next few years.

```{r}
dfmyocc |> summarize(
  .by = occupational_title,
  pred_jobs = sum(projected_employment)
) |> ggplot(aes(x=reorder(occupational_title,pred_jobs), y=pred_jobs))+geom_col()+coord_flip()+labs(x='Occupational Title', y='Total Predicted Jobs', title = 'Occupations by Predicted Jobs')
```

Surprisingly to me, `Secondary School Teacher` tops the list, with `Bartender` as a close second. I would have expected Bartender to be the highest, as it requires much less education and training, and is generally available everywhere.

Next, what skills are required for these occupations? Note that since I only have 10 occupations, I am measuring the skills by the number of predicted jobs.

```{r}
dfmyocc |> summarize(
  .by = skill_def,
  tot_jobs = sum(projected_employment),
) |> ggplot(aes(x=reorder(skill_def,tot_jobs), y=tot_jobs)) +geom_col() +coord_flip()+labs(x='Top Skill', y='Total Predicted Jobs', title = 'Top Skills by Predicted Jobs')
```

Considering half of my selected occupations are in the education sector, it isn't surprising that `Instructing` is the highest on this graph. Once again, we see `Active Listening` high up there. Ultimately, all of these skills are useful in the jobs I am looking at, and they are skills I believe I possess at high levels.

Next, lets look at education requirements. Again, these are based off the number of predicted jobs.

```{r}
dfmyocc |> summarize(
  .by = typical_education_requirement,
  tot_jobs = sum(projected_employment)
)|> ggplot(aes(x=reorder(typical_education_requirement,tot_jobs), y=tot_jobs)) + geom_col() + coord_flip()+labs(x='Education Requirement', y='Total Predicted Jobs', title = 'Minimum Education Requirement by Predicted Jobs')+scale_y_continuous(labels = label_number())
```

Unsurprisingly, `Bachelor's Degree` and `No Formal Educational Credential` top the charts. As with the last two graphs, the large amount of `Secondary School Teacher` and `Bartender` jobs skew the data. However, it is at least reassuring that the most jobs require a `Bachelor's Degree`, which is what I am working towards.

Most importantly, I would like to look at annual salary.

```{r}
dfmyocc |> summarize(
  .by = occupational_title,
  avg_sal = mean(average_annual_salary, na.rm=T)
) |> ggplot(aes(x=reorder(occupational_title, avg_sal), y=avg_sal))+geom_col()+coord_flip()+labs(x='Occupational Title', y='Average Annual Salary', title = 'Occupations by Annual Salary')
```

There aren't a whole lot of surprises here. `Actuary` tops the chart as expected, and `Bartender` is the lowest.

I do think that `Bartender` is a bit misleading, however. The hourly wage for bartenders is between \$11-14, which is fairly close to my base hourly wage of \$10 when I worked as a bartender. My assumption is that tips are not included in the data in this set, since with tips my hourly pay averaged around \$15-20 depending on the season.

The biggest surprise to me was `Art, Drama, & Music Teacher, Postsecondary` having the third highest salary. I would have expected Statistician to be higher at least.

Another surprise is how high `Tutor` is on the list. Upon further inspection of the data, `Tutor` is split into two groups. One group has an average hourly wage of around \$20, which is close to what I currently make as a tutor. The other group has an average hourly wage of around \$50, which pulls up the overall average salary quite a bit. So it looks like the biggest thing I have learned from this project is that I need to ask for a raise.

## Final Thoughts

Through this data exploration, I have gained some valuable insights into the job market in Iowa. There are a lot of takeaways from the initial general exploration, such as `Active Listening` being far and away the most widely sought skill. However, the analysis centered around my personal job market was much more impact to me.

I walk away from this project with a lot more confidence. Not only are the jobs that I am looking at plentiful, but their skills line up with skills I have built up over the years. It is also super relieving to know that financially the market doesn't meaningfully change based on location.
